---
title: "RNA-seq Differential Expression Report"
subtitle: "Breast Cancer Tumor vs Normal — GSE183947"
author: "Lingunja | CBU Bioinformatics"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
params:
  ref_condition:   "normal"
  treat_condition: "tumor"
  padj_cutoff:     0.05
  lfc_cutoff:      1.0
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
source("scripts/deseq2.R")

counts  <- load_counts("results/counts/counts.txt")
coldata <- build_coldata(counts, params$ref_condition, params$treat_condition)
dds     <- run_deseq2(counts, coldata)
results <- get_results(dds, params$ref_condition, params$treat_condition,
                       params$padj_cutoff, params$lfc_cutoff)
res_df  <- results$res_df
summary <- summarise_de(res_df, params$padj_cutoff, params$lfc_cutoff)
```

## 1. Summary

| Parameter | Value |
|---|---|
| Comparison | `r params$treat_condition` vs `r params$ref_condition` |
| Samples | `r ncol(counts)` |
| Genes tested | `r format(summary$total_tested, big.mark=",")` |
| Significant DEGs | **`r summary$total_sig`** |
| Upregulated | `r summary$up` |
| Downregulated | `r summary$down` |

## 2. PCA

```{r pca}
plot_pca(dds)
```

## 3. Volcano Plot

```{r volcano, fig.height=7}
plot_volcano(res_df, params$padj_cutoff, params$lfc_cutoff)
```

## 4. Heatmap

```{r heatmap, fig.height=8}
plot_heatmap(dds, res_df)
```

## 5. Top Upregulated Genes

```{r top_up}
knitr::kable(summary$top10_up, digits=c(0,3,5))
```

## 6. Top Downregulated Genes

```{r top_down}
knitr::kable(summary$top10_down, digits=c(0,3,5))
```

## 7. Full Results

```{r table}
DT::datatable(
  res_df %>% filter(!is.na(padj)) %>%
    select(gene, log2FoldChange, padj, direction),
  filter="top", rownames=FALSE,
  options=list(pageLength=15, scrollX=TRUE)
)
```

## 8. Methods

Reads were quality-assessed with **FastQC**, trimmed with **fastp**, aligned to **GRCh38** using **HISAT2**, and quantified with **featureCounts**. Differential expression was performed with **DESeq2** (LFC shrinkage via ashr). Significance threshold: FDR < `r params$padj_cutoff`, |log₂FC| > `r params$lfc_cutoff`.

---
*Lingunja | CBU Bioinformatics*
